---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# ethDate

<!-- badges: start -->
[![R-CMD-check](https://github.com/GutUrago/ethDate/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/GutUrago/ethDate/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/GutUrago/ethDate/graph/badge.svg)](https://app.codecov.io/gh/GutUrago/ethDate)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

This package provides a robust and efficient solution for converting between Ethiopian and Gregorian dates. Built with high-performance C++ code via `{Rcpp}`, it ensures lightning-fast computations. It has built-in checks for leap years and 13th month (Pagume).

## Key Features:

- **Seamless Date Conversion:** Effortlessly convert Ethiopian dates to and from Gregorian dates.
- **High-Speed Computation:** Optimized with {Rcpp} for maximum efficiency.
- **Date Arithmetic:** Perform date additions and subtractions with precision.
- **Day-Based Units:** Dates are represented as the number of days since a defined base date (1970-01-01).

🚀 **Upcoming Features:** Future versions will extend support for time and time zones.

## Installation

You can install the development version of ethDate like so:

``` r
# install.packages("devtools")
devtools::install_github("GutUrago/ethDate")
```


## Base Date

The default `base` package origin is '1970-01-01' and `ethDate` does the same with 
equivalent Ethiopian date.

```{r base-origin}
as.Date(0)
```

```{r ethDate-origion}
library(ethDate)
as_eth_date(0)
```

You can convert `ethDate` object to Gregorian calendar to get equivalent date. For 
the above example we can do:


```{r convert-to_gre}
to_gregorian(as_eth_date(0))
```

```{r convert-to_eth}
to_ethiopian(as.Date(0))
```

## Parse Ethiopian Dates

Ethiopian dates can be parsed from integers (number of days since 1970-01-01), 
character, and of course, Gregorian dates.

1. **Integers**

Use negative integers to represent date before the base date.

```{r from-int}
as_eth_date(20000)
```

```{r negative-int}
as_eth_date(-20000)
```


2. **Character**

```{r}
as_eth_date("2017-06-13")
```


```{r}
as_eth_date("2010/07/13", sep = "/")
```

```{r}
as_eth_date("06/13/2011", sep = "/", orders = "dmy")
```

3. **Date**

```{r}
as_eth_date(as.Date("2025-01-01"))
```


## Date Arithmetic

You can add or subtract scalar date to and from the `ethDate` object. The scalar represent number of days.

```{r}
date_1 <- as_eth_date("2017-01-01")
date_1 + 30
date_1 + 365
date_1 + 7
date_1 - 7
date_1 - 366
```




## Example

In this example, we use `lakers` data from `lubridate` package and convert dates 
back and forth two times to check it's consistency.

```{r warning=FALSE, message=FALSE}
library(tidyverse)

df <- lakers |> 
  mutate(date_gre = ymd(date),
         date_eth = as_eth_date(date_gre),
         date_gre2 = to_gregorian(date_eth),
         date_eth2 = to_ethiopian(date_gre2)) |> 
  select(starts_with("date"))


kableExtra::kable(head(df))
```


Let's confirm it's consistency:

```{r}
all(df$date_gre == df$date_gre2)
```

```{r}
all(df$date_eth == df$date_eth2)
```



```{=html}
<div>
<h1 style="text-align: center; color=blue">The End!</h1>
</div>
```






