---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# ethiodate

<!-- badges: start -->
[![R-CMD-check](https://github.com/GutUrago/ethiodate/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/GutUrago/ethiodate/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/GutUrago/ethiodate/graph/badge.svg)](https://app.codecov.io/gh/GutUrago/ethiodate)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

This package provides a robust and efficient solution for working with Ethiopian dates. It can seamlessly convert to and from Gregorian dates. It ensures lightning-fast computations thanks to the `{Rcpp}` package that enables the integration of high-performance C++ code. It has built-in checks for leap years and the 13th month (Pagume).

## Key Features:

- **Seamless Date Conversion:** Effortlessly convert Ethiopian dates to and from Gregorian dates.
- **High-Speed Computation:** Optimized with {Rcpp} for maximum efficiency.
- **Date Arithmetic:** Perform date additions and subtractions with precision.
- **Day-Based Units:** Dates are represented as the number of days since 1963-04-23 EC (1970-01-01 GC).

🚀 **Upcoming Features:** Future versions will extend support for time and time zones.

## Installation

You can install the development version of `ethiodate` package like so:

``` r
# install.packages("devtools")
devtools::install_github("GutUrago/ethiodate")
```


## Origin Date

The default `base` package origin date is '1970-01-01 GC' and `ethiodate` does the same with equivalent Ethiopian date, which is '1962-04-23 EC'. Both base `Date` object and `ethdate` internally stores number of days since origin date. The days before the origin date are represented with negative values. This makes computation easier.

```{r base-origin}
# The same origin as 'Date' object
library(ethiodate)
as.Date(0) == vctrs::vec_data(eth_date(0))

as.Date(0)
eth_date(0)
```

You can convert `ethdate` object to Gregorian calendar to get equivalent date. For the above example we can do:


```{r convert-to_gre}
as.Date(eth_date(0))
```

```{r convert-to_eth}
eth_date(as.Date(0))
```

## Parsing

Ethiopian dates can be parsed from integers (number of days since 1962-04-23 EC / 1970-01-01 GC), separate vector of year components, and of course, Gregorian dates.

1. **Integers**

Use negative integers to represent date before the origin date.

```{r from-int}
eth_date(20000)
```

```{r negative-int}
eth_date(-20000)
```


2. **Year components**

```{r}
eth_make_date(2017, 01, 01)
```

3. **Date**

```{r}
eth_date(as.Date("2025-01-01"))
```


## Date Arithmetic

You can add or subtract scalar date to and from the `ethDate` object. The scalar represent number of days.

```{r}
date_1 <- eth_make_date(2017,01,01)
date_1 + 400
date_1 - 1
```

You can subtract date object as well.

```{r}
eth_make_date(2017, 01, 25) - eth_make_date(2017, 01, 10)

eth_date(0) - eth_date(as.Date("1970-01-01"))

```

## Formatting

You can nicely format Ethiopian like you do with base `Date` object. This makes it easy to use with inline code for reports and exporting files.

```{r formatting}
x <- eth_make_date(2000, 1, 15)
format(x, format = "%B %d, %Y")
```
```{r more-format}
format(x, format = "%B %d, %Y", lang = "lat")
```


```{r systime}
format(eth_date(Sys.Date()), format = "This file was updated on %B %d, %Y EC.", lang = "en")
```


## Example

In this example, we use `lakers` data from `lubridate` package and convert dates back and forth two times to check it's consistency.

```{r warning=FALSE, message=FALSE}
library(tidyverse)

df <- lakers |> 
  mutate(date_gre = ymd(date),
         date_eth = eth_date(date_gre),
         date_gre2 = as.Date(date_eth),
         date_eth2 = eth_date(date_gre2)) |> 
  select(starts_with("date"))


kableExtra::kable(sample_n(df, 5))
```


Let's confirm it's consistency:

```{r}
all(df$date_gre == df$date_gre2)
```

```{r}
all(df$date_eth == df$date_eth2)
```

That proves it consistently converted dates back and forth for all **34624** observations in the `lakers` dataset.


```{=html}
<div>
<h1 style="text-align: center; color=blue">The End!</h1>
</div>
```

