---
title: "Introduction to ethiodate"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ethiodate}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


The Ethiopian calendar is widely used in Ethiopia and differs significantly from the Gregorian calendar. It has 13 months: 12 months of 30 days and a 13th month (Pagumē) of 5 or 6 days, depending on the year.

The `ethiodate` package provides tools for parsing, converting, and working with Ethiopian dates in R. This vignette demonstrates how to parse Ethiopian dates from strings or numeric formats and convert them into `ethdate` objects.


# Parsing Ethiopian Dates

## From Character Strings

You can parse Ethiopian dates from standard character representations like "2015-01-01" (Ethiopian calendar). Use `eth_parse_date()`:


```{r setup}
library(ethiodate)
eth_parse_date("2015-01-01")
```

You can also provide a vector of dates:


```{r}
eth_dates <- c("2015-01-01", "2015-02-15", "2015-13-05")
eth_parse_date(eth_dates)
```

If you are parsing dates from string, you must provide correct formatting using `format` argument and the string must have consistent pattern. This argument accepts ISO 8601 formats, use `?strptime` for more.

***Examples***

```{r}
x <- c("01/17/2025", "05/12/2017")
eth_parse_date(x, format = "%m/%d/%Y")
```

You can also parse date that contains month names if specify the exact format as follows:

```{r}
eth_parse_date("Meskerem 25, 2017", format = "%B %d, %Y")
eth_parse_date("መስከረም 25, 2017", format = "%B %d, %Y", lang = "amh")
eth_parse_date("Sep 25, 2017", format = "%b %d, %Y", lang = "en")
```


## From Numeric

`ethate` is just a numeric vector, under the hood, representing number of days since 1970-01-01 GC (1962-04-12 EC). The date before the origin is represented by negative values. This makes it to seamlessly integrate features if base R Date object. For instance;


```{r}
# The same origin
eth_parse_date("1962-04-23") |> unclass()
as.Date("1970-01-01") |> unclass()
```

So, you can supply number of days from a custom origin to construct Ethiopian date object. If the origin is missing, the default is 1962-04-23 EC. See `?eth_date` for more.

```{r}
eth_date(0, origin = lubridate::today())
```

## From Components

Constructing `ethdate` object from separate vector of year, month and day is supported.

```{r}
eth_make_date(2017, 1, 1)

y <- c(2022, 2025)
m <- c(5, 9)
d <- c(15, 19)
eth_make_date(y, m, d)

```

# Invalid and Missing Values

`ethiodate` validates every values before coercing it to `ethdate` object. For instance, 13th month has 6 days only during leap year. If you supply such erroneous values, you will `NA's` along with warnig message.

```{r}
# 2011 is leap year -> Correct
eth_date("2011-13-6")

# Incorrect
eth_date("2012-13-6")
```

# Conversion

Conversion between Ethiopian and Gregorian calendar is fully supported. 

***To Ethiopian***

```{r}
gre_date <- as.Date("2025-01-15")
eth_date(gre_date)
```

***To Gregorian***

```{r}
eth_dt <- eth_date(0)
as.Date(eth_dt)
```


# Operations

As any other date objects, `ethdate` supports date operations. A unit any arithmetric operations is date. 

```{r}
# Adding scalar days
eth_date("2010-09-14") + 6

# Subtraction
eth_date("2010-09-14") - 6

# Differences
eth_date("2010-09-14") - eth_date("2010-09-10")

```

# Formatting

You can format `ethdate` object like any other date objects. 

```{r}
format(eth_today(), format = "This documents was updated on %B %d, %Y EC.")
```


# Conclusion

Parsing Ethiopian dates is a crucial step in analyzing data collected in the Ethiopian calendar system. The `ethiodate` package provides a consistent and robust interface for converting string representations of Ethiopian dates into properly classed objects ready for further analysis or conversion.

For more functionality, see `?eth_date`, `?eth_parse_date`, `?eth_make_date`, and the rest of the package documentation.



